Deep Reinforcement Learning for Conservation Decisions

### 2021-04-28 manuscript/manuscript.Rmd

---
title: Deep Reinforcement Learning for Conservation Decisions
authors:
  - name: Marcus Lapeyrolerie
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
  - name: Melissa Chapman
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
  - name: Kari Norman
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
  - name: Carl Boettiger
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
    email: cboettig@berkeley.edu

abstract: |
  Enter the text of your abstract here.
keywords:
  - blah
  - blee
  - bloo
  - these are optional and can be removed
bibliography: references.bib
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}

output: 
  rticles::arxiv_article:
    keep_tex: true
---

@@

Deep Reinforcement Learning (RL), a machine learning technique in which a _software agent_ is trained to select actions which best maximize some cumulative reward, could hold significant untapped promise for solving complex ecological and conservation decision-making problems from wildfire management to biodiversity preservation to zoonotic pandemics.
Unlike more widely recognized machine learning techniques for either _supervised_ learning (such as image classification, @species_ex, @satellite_ex) or unsupervised learning, RL focuses on solving _dynamic_ problems which require planning ahead, often in uncertain and changing environments.
Of all machine learning techniques, RL most resembles a general artificial intelligence.
Notable successes using RL include AlphaZero, in which a software agent learned to beat the world's best human players in games of chess, Go, and Shogi [@alphazero],
and major advances in robotics, including decision-making in autonomous vehicles [@robotics_ex]. 
These tasks all require the agent to not only anticipate future states of it's environment, but also respond to those changes to influence future outcomes.

<!-- Millie help on this para? -->
If conservation challenges are typically complex and dynamics, existing approaches are largely simple and static.
The establishment of protected areas (cite MPAs, 30x30, etc) focuses on policies fixed in space and time to address challenges that are fixed in neither.
The endangered species act ...
Ecosystem services ....
Such simplicity is in no small part due to the challenge associated in determining a 
Some reference to @TomDietrich on potential for CS-based application.

In contrast to other ML methods, RL does not require massive data sets for training.  
Despite the expansion in data collection from remote sensors, citizen science, and other efforts, ecological data remains very sparse on many necessary details.
Moreover, no amount of data can train supervised or unsupervised ML approaches can predict a future which has no analog in the historical record.
Mechanistic, process-based models remain our most powerful tool to explore scenarios of dynamic ecosystems in changing climates.
Methods from _optimal control theory_ have long been used in conservation and resource economics literature [@Clark1990] to determine optimal policies given such models [@Polasky2011].
However, control theory approaches are also tightly constrained by the "curse of dimensionality" [@Mangel1980; @Marescot2013], limiting their application to only simple, stylized models.
Deep reinforcement learning offers a way forward. 
Deep RL agents are trained not with mountains of data, but by interacting with simulators.
Most recent advances in Deep RL have illustrated algorithms trained by playing arcade game simulators such as Atari BreakOut, or physics-based simulations of robotics movement [@TD3; @SAC; @PPO; @A2C].
Simulating from complex models is far simpler than solving them.
The richest ecosystem models, such as the global circulation models (GCMs) used to generate future climate scenarios [@something], or individual-based models used to predict forest growth [@something] or wildfire spread [@something], are massive simulations. 
Given sufficient computational power, RL can be trained and evaluated across a broad range of possible realistic scenarios well beyond the reach of classical methods.
In this manner, RL would compliment, rather than replace, existing empirical and theoretical research which would continue to form the basis for those simulations.

If RL has great promise, the approach also imposes great risk.  
Lacking the tractability of classical methods, an RL agent can appear to solve a wide class of scenarios with ease, but may fail unexpectedly under seemingly similar conditions [@henderson2017].  
Despite recent progress, Deep RL is a young field and stability and robustness of its methods remain largely open problems. 
RL approaches are no panacea, and improper design or testing can easily lead to negative outcomes.
Technical issues are not the only pitfall of AI applications to conservation problems.
Giving algorithms direct influence over conservation policy may also raise ethical and political issues, particularly if those algorithms (or components thereof) are treated as proprietary intellectual property.

@@

@@

<!-- meh, not sure we need this? -->
Reinforcement learning focuses primarily on sequential decision problems: tasks in which a decision-maker or "agent" must repeatedly decide what action to take in response to new information about the "state" of the system they seek to manage, in order to maximize some objective (referred to  as the reward or utility).
Any conservation problem that considers the future possible scenarios is a sequential decision problem: even scenarios that face a one-off decision such as whether declare some region as a protected area, or list or de-list a species as endangered, are sequential decision problems if they must consider the timing of their actions.
In contrast, decisions that do not consider the future, such as deciding which of a collection of possible regions should be prioritized based only on the present: how many species currently inhabit a region, how much each region would cost to purchase today, are not sequential problems.
Central to any sequential decision problem is the need to consider possible future states, the need to forecast.
Because the agent's actions will influence that future, agents facing sequential decision problems must further be able to forecast how each possible action available to them will change that forecast.
Consequentially, our lack of knowledge about the underlying models and inability to predict future environments in our rapidly changing world pose a fundamental challenge to conservation decision-making.

@@

Before providing some examples on how deep reinforcement learning can be applied to ecological and conservation problems, we will provide a brief, general introduction to RL.
The basic problem set-up of RL is as follows: at a time step, $t$, the agent observes a representation of the environment's state, $s_t$, and must accordingly select an action, $a_t$; at the next time step, the agent will observe a new state, $s_{t+1}$, and will receive a numerical reward, $r_{t+1}$.
This new state, $s_{t+1}$, and reward, $r_{t+1}$, are determined by the environment.
The task for the agent is to learn a _policy_, $\pi$, that prescribes what action the agent should perform after observing a certain state.
The agent interacts with the environment via trial and error in what is called the _training_ period, during which the agent continally updates its policy in order to maximize its expected cumulative sum of rewards.

@@

\[
  p_\pi(\tau) = d_0(s_0) \prod_{t=0}^H \pi(a_t \mid s_t) T(s_{t+1} \mid s_t, a_t)
\]

@@

\[
  \pi^* = \underset{\pi}{\text{argmax}} \,\, \mathbb{E}_{\tau \sim p_\pi(\tau)}\Big[\sum_{t=0}^{H} \gamma^t r(s_t, a_t) \Big]
\]

@@

\[
  V^\pi(s_t) = \mathbb{E}_{\tau \sim p_\pi(\tau \mid s_t)}\Bigg[\sum_{t' = t}^H \gamma^{t' - t} r(s_t, a_t)\Bigg]
\]

Meanwhile, the state-action-value or quality function, finds the expected return given a state and action,
\[
  Q^\pi(s_t, a_t) = \mathbb{E}_{\tau \sim p_\pi(\tau \mid s_t, a_t)}\Bigg[\sum_{t' = t}^H \gamma^{t' - t} r(s_t, a_t)\Bigg]
\]

And lastly, the advantage function characterizes a relative state-action value in that it quantifies how valuable a state-action pair is over a baseline. The advantage function can then be succintly defined as
\[
  A^\pi(s_t, a_t) = Q^\pi(s_t, a_t) - V^\pi(s_t)
\]
Learning $Q^\pi$, commonly called the Q-value function, is more powerful than learning the state-value function as Q-values easily prescribe what action to take when in a certain state, $a_t^* = \underset{a_t}{\text{argmax}} \, Q^\pi (s_t, a_t)$. 
Conversely, if we know the state-value function, we would have to perform a maximization over an expected value to find the optimal action, $a_t^* = \underset{a_t}{\text{argmax}} \, \mathbb{E}_{s_{t+1} \sim \,T(s_{t+1} \mid s_t, a_t)}[V^\pi(s_{t+1})]$, which is less tractable.
Yet, learning the advantage function is often preferred in practice as advantage estimators favorably have lower variance than state-value and Q-value estimators [cite schulman gae].

@@

<!--

@@

-->

There is an ever increasing multitude of reinforcement learning algorithms, and while there are far too many algorithms to discuss in detail in this paper, we will herein classify and touch on some current methods. 
RL algorithms can generally be split into 2 groups, model-based and model-free RL.
In model-based RL, the agent is attempting to learn the state transition operator, $T$.
Once the transition model is known, the agent can then use the transition model to plan out what is the optimal action, often with Model Predictive Control [cite tassa 2012 MPC], or the agent can use the transition model to generate more samples to aid learning [cite Sutton dyna, alphago?].
In model-free RL, the agent does not attempt to learn a dynamics model and instead tries to either learn a parameterized policy, learn a value function, or try to learn both a parameterized policy and a value function. The general trade-off between model-free and model-based approaches is that in practice, model-based algorithms are often more sample efficient than model-free algorithms but often fail to surpass model-free algorithms in overall agent performance [cite?].

@@

```{=latex}
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
Algorithm & Description & Model & Policy & Action Space & State Space \\
\hline
MBPO & Model-based Policy Optimization & Model-based & On-policy& &\\
\hline
MCTS & Monte Carlo Tree Search & Model-based & Either & & \\
\hline
A2C  & Advantage Actor Critic & Model-free & On-policy& &\\
\hline
A3C  & Asynchronous A2C & Model-free & On-policy & &\\
\hline
TRPO  & Trust Region Policy Optimization & Model-free & On-policy & &\\
\hline
PPO  & Proximal Policy Optimization & Model-free & On-policy & &\\
\hline
DQN & Deep Q Networks & Model-free & Off-policy& &\\
\hline
DDPG & Deep Deterministic Policy Gradient & Model-free & Off-policy & &\\
\hline
TD3 & Twin Delayed DDPG & Model-free & Off-policy & &\\
\hline
SAC & Soft Actor Critic & Model-free & Off-policy & &\\
\hline
IMPALA & Importance Weighted Actor Learner & Model-free & Off-policy & &\\
\hline
APE-X & Distributed Prioritized Experience Replay & Model-free & Off-policy & & \\
\hline
\end{tabular}
\end{center}
```

# Examples

@@

This provides a natural first benchmark for deep RL approaches, since we can compare the RL solution to the mathematical optimum directly.
The problem is also of interest because the RL approach can be extended to consider ecological dynamics that either surpass the complexity possible with SDP methods, or violate assumptions of such classical approaches).
Moreover, fisheries management is both an important global challenge in it's own right [@Worm2006; @Costello2018] as well as a common test case to understand ecological dynamics and conservation management more broadly [@].

For illustrative purposes, we consider the simplest version of the dynamic optimal harvest problem as outlined by @Clark1974 (for the deterministic case) and @Reed1979 (under stochastic recruitment).
In short, the manager seeks to optimize the net present value (discounted cumulative catch) of a fishery, observing the stock size each year and setting an appropriate harvest quota in response.
In the classical approach, the best model of the fish population dynamics must first be estimated from data, potentially with posterior distributions over parameter estimates reflecting any uncertainty.
From this model, the optimal harvest policy -- that is, the a function which returns the optimal quota for each possible observed stock size -- can be determined either by analytic [@Reed1979] or numerical [@Marescot2013] methods, depending on the complexity of the model.
In a RL approach, the model is instead replaced by a computer simulation, usually referred to as an "environment."
The simplest simulation is merely an implementation of the estimated model itself.
However, unlike an analytic or SDP solution that is derived directly from the estimated model, the RL approach makes no assumption about the precise functional form or parameter values underlying the dynamics -- it is in principle agnostic to the details of the simulation.

To better illustrate the potential differences between different RL algorithms used to train the agents, we will examine two RL solutions, one based on A2C [@A2C] and the other based on @TD3.
A step-by-step walk-through for training both agents on this environment is provided in the Appendix.
We compare the resulting management, policy, and reward under either RL agent to that achieved by the optimal management solution [Fig 1].
Despite having no knowledge of the underlying model and clear differences in the resulting strategies, both RL agents learn enough to achieve nearly optimal performance.

@@

Our second example focuses on a case for which we do not already have an existing, provably optimal policy to compare against.
We consider the generic problem of an ecosystem facing slowly deteriorating environmental conditions which move the dynamics closer towards a tipping point [Fig 2].
This model of a critical transition has been posited widely in ecological systems, from the simple consumer-resource model of @May1977 on which our dynamics are based, to microbial dynamics [@Dai2013] lake ecosystem communities [@Carpenter2008] to planetary ecosystems [@Barnkosky2014].
On top of these ecological dynamics we introduce an explicit ecosystem service model quantifying the value of more desirable 'high' state relative to the 'low' state.
For simplicity, we assume a proportional benefit $b$ associated with the ecosystem state $X(t)$.
Thus when the ecosystem is near the 'high' equilibrium $\hat X_H$, the corresponding ecosystem benefit $b \hat X_H$ is higher than at the low equilibrium, $b x_L$, consistent with the intuitive description of ecosystem tipping points [@Barnosky2014]. 
We also enumerate the possible actions which a manager may take in response to environmental degradation. 
In the absence of any management response, we assume the environment deteriorates at a fixed rate $\alpha$, which can be thought of as the incremental increase in global mean temperature or similar anthropogenic forcing term.
Management can respond to slow or even reverse this trend by choosing an opposing action $A_t$.
We assume that large actions are proportionally more costly than small actions, consistent with the expectation of diminishing returns: taking the cost associated with an action $A_t$ as equal to $c A_t^2$.
Many alterations of these basic assumptions are also possible: our `gym_conservation` implements a range of different scenarios with user-configurable settings.
In each case, the manager observes the current state of the system each year and must then select the policy response that year.

@@

Because this problem involves a parameter whose value changes over time (the slowly deteriorating environment), the resulting ecosystem dynamics are not autonomous.
This precludes our ability to solve for the optimal management policy using classical theory such as for Markov Decision Processes (MDP, @Marescot2013), typically used to solve sequential decision-making problems.
However, it is often argued that simple rules can achieve nearly optimal management of ecological conservation objectives in many cases [@Possingham; @Possingham; @morePossingham].
A common conservation strategy employs a fixed response level rather than a dynamic policy which is toggled up or down each year: for example, declaring certain regions as protected areas in perpetuity.
An intuitive strategy faced with an ecosystem tipping point would be 'perfect conservation', in which the management response is perfectly calibrated to counter-balance any further decline.
While the precise rate of such decline may not be known in practice (and will not be known to RL algorithms before-hand either), it is easy to implement such a policy in simulation for comparative purposes. 
We compare this rule-of-thumb to the optimal policy found by training an agent using the TD3 algorithm.

@@

Examining the replicate management trajectories and corresponding rewards [Fig 3] reveal that the RL agent incurs significantly higher costs in the initial phases of the simulation, dipping the cumulative reward into negative territory on average and well below the reward realized by the simple steady-state strategy.
This initial investment then begins to pay off -- by about 200th time step the RL agent has surpassed the performance of the steady state strategy.  
The policy plot provides more intuition for the RL agent's strategy: at very high state values, the RL agent opts for no conservation action -- so far from the tipping point, no response is required.
Near the tipping point, the RL agent steeply ramps up the conservation effort, and retains this effort even as the system falls below the critical threshold, where a sufficiently aggressive response can tip the system back into recovery.
For a system at or very close to the zero-state, the RL agent gives up, opting for no action.
Recall that the quadratic scaling of cost makes the rapid response of the TD3 algorithm much more costly to achieve the same net environmental improvement divided into smaller increments over a longer timeline.
However, our RL agent has discovered that the extra investment for a rapid response is well justified as the risk of crossing a tipping point increases.

@@

Such examination makes it clear how the RL agent's use of a more dynamic strategy out-preforms the steady-state strategy.
The various kinks in the RL policy function also suggest that this solution is nevertheless surely not the optimal solution -- such features are typical of RL-based policies, where such spandral features from early experimentation may become baked into the long-term strategy if adjusting them has little improvement on the final reward.
Further tuning of hyper-parameters, increased training, alterations to the training algorithm or an entirely different training algorithm would likely be able to further improve upon this performance.
The use of a recurrent neural network architecture instead of the multi-layer perceptron used here would open the door for the policy to reflect the non-autonomous dynamics more directly.
A recurrent neural network, such as long-short-term-memory (LSTM), could learn a strategy which reflects not only the current system state, but also latent information such as the time elapsed.

@@

-->

<!-- Something about relevance to both the _theory_ and practice of ecological conservation. (e.g. "strategic" 1-D models have probably had a greater influence, (for better or worse!) on real world policies than complex, detail-specific, precise "tactical" models) -->

@@

The examples presented here only scrape the surface of possible RL applicaations to conservation problems.
The examples we have focused on are intentionally quite simple, though it is worth remembering that these very same simple models have a long history of relevance and application in both research and policy contexts.
Despite their simplicity, the optimal strategy is not always obvious before hand, however intutive it may appear in retrospect.
In the case of the ecosystem tipping point scenario, the optimal strategy is unknown, and the approximate solution found by our RL implementation could almost certainly be improved upon.

@@

- wild west of AI?
-can separate RL from having gyms implemented -> how do you compare different models, 
-mention that customization is probably the way to go, baselines only go so far

>>>>

---
title: Deep Reinforcement Learning for Conservation Decisions
authors:

  - name: Marcus Lapeyrolerie
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
  - name: Melissa Chapman
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
  - name: Kari Norman
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
  - name: Carl Boettiger
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
    email: cboettig@berkeley.edu
abstract: |
  Enter the text of your abstract here.
keywords:
  - blah
  - blee
  - bloo
  - these are optional and can be removed
bibliography: references.bib
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
output: 
  rticles::arxiv_article:
    keep_tex: true
editor_options: 
  markdown: 
    wrap: sentence
---

@@

Deep Reinforcement Learning (RL), a machine learning technique in which a *software agent* is trained to select actions which best maximize some cumulative reward, could hold significant untapped promise for solving complex ecological and conservation decision-making problems from wildfire management to biodiversity preservation to zoonotic pandemics.
Unlike more widely recognized machine learning techniques for either *supervised* learning (such as image classification, [@species_ex], [@satellite_ex]) or unsupervised learning, RL focuses on solving *dynamic* problems which require planning ahead, often in uncertain and changing environments.
Of all machine learning techniques, RL most resembles a general artificial intelligence.
Notable successes using RL include AlphaZero, in which a software agent learned to beat the world's best human players in games of chess, Go, and Shogi [@alphazero], and major advances in robotics, including decision-making in autonomous vehicles [@robotics_ex].
These tasks all require the agent to not only anticipate future states of it's environment, but also respond to those changes to influence future outcomes.

<!-- Millie help on this para? -->

If conservation challenges are typically complex and dynamics, existing approaches are largely simple and static.
The establishment of protected areas (cite MPAs, 30x30, etc) focuses on policies fixed in space and time to address challenges that are fixed in neither.
The endangered species act ... Ecosystem services ....
Such simplicity is in no small part due to the challenge associated in determining a Some reference to [@TomDietrich] on potential for CS-based application.

In contrast to other ML methods, RL does not require massive data sets for training.\
Despite the expansion in data collection from remote sensors, citizen science, and other efforts, ecological data remains very sparse on many necessary details.
Moreover, no amount of data can train supervised or unsupervised ML approaches can predict a future which has no analog in the historical record.
Mechanistic, process-based models remain our most powerful tool to explore scenarios of dynamic ecosystems in changing climates.
Methods from *optimal control theory* have long been used in conservation and resource economics literature [@Clark1990] to determine optimal policies given such models [@Polasky2011].
However, control theory approaches are also tightly constrained by the "curse of dimensionality" [@Mangel1980; @Marescot2013], limiting their application to only simple, stylized models.
Deep reinforcement learning offers a way forward.
Deep RL agents are trained not with mountains of data, but by interacting with simulators.
Most recent advances in Deep RL have illustrated algorithms trained by playing arcade game simulators such as Atari BreakOut, or physics-based simulations of robotics movement [@TD3; @SAC; @PPO; @A2C].
Simulating from complex models is far simpler than solving them.
The richest ecosystem models, such as the global circulation models (GCMs) used to generate future climate scenarios [@something], or individual-based models used to predict forest growth [@something] or wildfire spread [@something], are massive simulations.
Given sufficient computational power, RL can be trained and evaluated across a broad range of possible realistic scenarios well beyond the reach of classical methods.
In this manner, RL would compliment, rather than replace, existing empirical and theoretical research which would continue to form the basis for those simulations.

If RL has great promise, the approach also imposes great risk.\
Lacking the tractability of classical methods, an RL agent can appear to solve a wide class of scenarios with ease, but may fail unexpectedly under seemingly similar conditions [@henderson2017].\
Despite recent progress, Deep RL is a young field and stability and robustness of its methods remain largely open problems
. RL approaches are no panacea, and improper design or testing can easily lead to negative outcomes
. Technical issues are not the only pitfall of AI applications to conservation problems
. Giving algorithms direct influence over conservation policy may also raise ethical and political issues, particularly if those algorithms (or components thereof) are treated as proprietary intellectual property
.

@@

<!-- Kari: We reference Markov Decision processes as an alternative approach in the tipping points example and just generally elsewhere so it might be helpful to talk about them in the introduction-->

#### Key Terminology

-   model (underlying ecological model)

```{=html}
<!-- -->
```
-   agent

-   trained agent (model)

-   environment

-   predict

-   policy

-   model-free

-   model-based

@@

<!-- meh, not sure we need this? -->

Reinforcement learning focuses primarily on sequential decision problems: tasks in which a decision-maker or "agent" must repeatedly decide what action to take in response to new information about the "state" of the system they seek to manage, in order to maximize some objective (referred to as the reward or utility).
Any conservation problem that considers the future possible scenarios is a sequential decision problem: even scenarios that face a one-off decision such as whether declare some region as a protected area, or list or de-list a species as endangered, are sequential decision problems if they must consider the timing of their actions.
In contrast, decisions that do not consider the future, such as deciding which of a collection of possible regions should be prioritized based only on the present: how many species currently inhabit a region, how much each region would cost to purchase today, are not sequential problems.
Central to any sequential decision problem is the need to consider possible future states, the need to forecast.
Because the agent's actions will influence that future, agents facing sequential decision problems must further be able to forecast how each possible action available to them will change that forecast.
Consequentially, our lack of knowledge about the underlying models and inability to predict future environments in our rapidly changing world pose a fundamental challenge to conservation decision-making.

@@

Before providing some examples on how deep reinforcement learning can be applied to ecological and conservation problems, we will provide a brief, general introduction to RL.
The basic problem set-up of RL is as follows: at a time step, $t$, the agent observes a representation of the environment's state, $s_t$, and must accordingly select an action, $a_t$; at the next time step, the agent will observe a new state, $s_{t+1}$, and will receive a numerical reward, $r_{t+1}$.
This new state, $s_{t+1}$, and reward, $r_{t+1}$, are determined by the environment.
The task for the agent is to learn a *policy*, $\pi$, that prescribes what action the agent should perform after observing a certain state.
The agent interacts with the environment via trial and error in what is called the *training* period, during which the agent continally updates its policy in order to maximize its expected cumulative sum of rewards.

@@

$$
  p_\pi(\tau) = d_0(s_0) \prod_{t=0}^H \pi(a_t \mid s_t) T(s_{t+1} \mid s_t, a_t)
$$

@@

$$
  \pi^* = \underset{\pi}{\text{argmax}} \,\, \mathbb{E}_{\tau \sim p_\pi(\tau)}\Big[\sum_{t=0}^{H} \gamma^t r(s_t, a_t) \Big]
$$

@@

$$
  V^\pi(s_t) = \mathbb{E}_{\tau \sim p_\pi(\tau \mid s_t)}\Bigg[\sum_{t' = t}^H \gamma^{t' - t} r(s_t, a_t)\Bigg]
$$

Meanwhile, the state-action-value or quality function, finds the expected return given a state and action, $$
  Q^\pi(s_t, a_t) = \mathbb{E}_{\tau \sim p_\pi(\tau \mid s_t, a_t)}\Bigg[\sum_{t' = t}^H \gamma^{t' - t} r(s_t, a_t)\Bigg]
$$

And lastly, the advantage function characterizes a relative state-action value in that it quantifies how valuable a state-action pair is over a baseline.
The advantage function can then be succintly defined as $$
  A^\pi(s_t, a_t) = Q^\pi(s_t, a_t) - V^\pi(s_t)
$$ Learning $Q^\pi$, commonly called the Q-value function, is more powerful than learning the state-value function as Q-values easily prescribe what action to take when in a certain state, $a_t^* = \underset{a_t}{\text{argmax}} \, Q^\pi (s_t, a_t)$.
Conversely, if we know the state-value function, we would have to perform a maximization over an expected value to find the optimal action, $a_t^* = \underset{a_t}{\text{argmax}} \, \mathbb{E}_{s_{t+1} \sim \,T(s_{t+1} \mid s_t, a_t)}[V^\pi(s_{t+1})]$, which is less tractable.
Yet, learning the advantage function is often preferred in practice as advantage estimators favorably have lower variance than state-value and Q-value estimators [cite schulman gae].

@@

```{=html}
<!--

@@

-->
```
There is an ever increasing multitude of reinforcement learning algorithms, and while there are far too many algorithms to discuss in detail in this paper, we will herein classify and touch on some current methods.
RL algorithms can generally be split into 2 groups, model-based and model-free RL.
In model-based RL, the agent is attempting to learn the state transition operator, $T$.
Once the transition model is known, the agent can then use the transition model to plan out what is the optimal action, often with Model Predictive Control [cite tassa 2012 MPC], or the agent can use the transition model to generate more samples to aid learning [cite Sutton dyna, alphago?]
.
In model-free RL, the agent does not attempt to learn a dynamics model and instead tries to either learn a parameterized policy, learn a value function, or try to learn both a parameterized policy and a value function.
The general trade-off between model-free and model-based approaches is that in practice, model-based algorithms are often more sample efficient than model-free algorithms but often fail to surpass model-free algorithms in overall agent performance [cite?]
.

@@

<!-- Kari: I think we need more background on what a neural network is, how it fits in, and what other options are - maybe use examples of hyperparameters that would be in all RL approaches not just those that use neural nets -->

```{=latex}
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
Algorithm & Description & Model & Policy & Action Space & State Space \\
\hline
MBPO & Model-based Policy Optimization & Model-based & On-policy& &\\
\hline
MCTS & Monte Carlo Tree Search & Model-based & Either & & \\
\hline
A2C  & Advantage Actor Critic & Model-free & On-policy& &\\
\hline
A3C  & Asynchronous A2C & Model-free & On-policy & &\\
\hline
TRPO  & Trust Region Policy Optimization & Model-free & On-policy & &\\
\hline
PPO  & Proximal Policy Optimization & Model-free & On-policy & &\\
\hline
DQN & Deep Q Networks & Model-free & Off-policy& &\\
\hline
DDPG & Deep Deterministic Policy Gradient & Model-free & Off-policy & &\\
\hline
TD3 & Twin Delayed DDPG & Model-free & Off-policy & &\\
\hline
SAC & Soft Actor Critic & Model-free & Off-policy & &\\
\hline
IMPALA & Importance Weighted Actor Learner & Model-free & Off-policy & &\\
\hline
APE-X & Distributed Prioritized Experience Replay & Model-free & Off-policy & & \\
\hline
\end{tabular}
\end{center}
```
# Examples

@@

This provides a natural first benchmark for deep RL approaches, since we can compare the RL solution to the mathematical optimum directly.
The problem is also of interest because the RL approach can be extended to consider ecological dynamics that either surpass the complexity possible with SDP methods, or violate assumptions of such classical approaches).
Moreover, fisheries management is both an important global challenge in it's own right [@Worm2006; @Costello2018] as well as a common test case to understand ecological dynamics and conservation management more broadly [\@].

For illustrative purposes, we consider the simplest version of the dynamic optimal harvest problem as outlined by [@Clark1974] (for the deterministic case) and [@Reed1979] (under stochastic recruitment).
In short, the manager seeks to optimize the net present value (discounted cumulative catch) of a fishery, observing the stock size each year and setting an appropriate harvest quota in response.
In the classical approach, the best model of the fish population dynamics must first be estimated from data, potentially with posterior distributions over parameter estimates reflecting any uncertainty.
From this model, the optimal harvest policy -- that is, the a function which returns the optimal quota for each possible observed stock size -- can be determined either by analytic [@Reed1979] or numerical [@Marescot2013] methods, depending on the complexity of the model.
In a RL approach, the model is instead replaced by a computer simulation, usually referred to as an "environment." The simplest simulation is merely an implementation of the estimated model itself.
However, unlike an analytic or SDP solution that is derived directly from the estimated model, the RL approach makes no assumption about the precise functional form or parameter values underlying the dynamics -- it is in principle agnostic to the details of the simulation.

To better illustrate the potential differences between different RL algorithms used to train the agents, we will examine two RL solutions, one based on A2C [@A2C] and the other based on [@TD3] <!-- Kari: are these different algorithms?-->.
A step-by-step walk-through for training both agents on this environment is provided in the Appendix.
We compare the resulting management, policy, and reward under either RL agent to that achieved by the optimal management solution [Fig 1].
Despite having no knowledge of the underlying model and clear differences in the resulting strategies, both RL agents learn enough to achieve nearly optimal performance.

@@

Our second example focuses on a case for which we do not already have an existing, provably optimal policy to compare against.
We consider the generic problem of an ecosystem facing slowly deteriorating environmental conditions which move the dynamics closer towards a tipping point [Fig 2].
This model of a critical transition has been posited widely in ecological systems, from the simple consumer-resource model of [@May1977] on which our dynamics are based, to microbial dynamics [@Dai2013] lake ecosystem communities [@Carpenter2008] to planetary ecosystems [@Barnkosky2014].
On top of these ecological dynamics we introduce an explicit ecosystem service model quantifying the value of more desirable 'high' state relative to the 'low' state.
For simplicity, we assume a proportional benefit $b$ associated with the ecosystem state $X(t)$.
Thus when the ecosystem is near the 'high' equilibrium $\hat X_H$, the corresponding ecosystem benefit $b \hat X_H$ is higher than at the low equilibrium, $b x_L$, consistent with the intuitive description of ecosystem tipping points [@Barnosky2014].
We also enumerate the possible actions which a manager may take in response to environmental degradation.
In the absence of any management response, we assume the environment deteriorates at a fixed rate $\alpha$, which can be thought of as the incremental increase in global mean temperature or similar anthropogenic forcing term.
Management can respond to slow or even reverse this trend by choosing an opposing action $A_t$.
We assume that large actions are proportionally more costly than small actions, consistent with the expectation of diminishing returns: taking the cost associated with an action $A_t$ as equal to $c A_t^2$.
Many alterations of these basic assumptions are also possible: our `gym_conservation` implements a range of different scenarios with user-configurable settings.
In each case, the manager observes the current state of the system each year and must then select the policy response that year.

@@

Because this problem involves a parameter whose value changes over time (the slowly deteriorating environment), the resulting ecosystem dynamics are not autonomous.
This precludes our ability to solve for the optimal management policy using classical theory such as for Markov Decision Processes (MDP, [@Marescot2013]), typically used to solve sequential decision-making problems.
However, it is often argued that simple rules can achieve nearly optimal management of ecological conservation objectives in many cases [@Possingham; @Possingham; @morePossingham].
A common conservation strategy employs a fixed response level rather than a dynamic policy which is toggled up or down each year: for example, declaring certain regions as protected areas in perpetuity.
An intuitive strategy faced with an ecosystem tipping point would be 'perfect conservation', in which the management response is perfectly calibrated to counter-balance any further decline.
While the precise rate of such decline may not be known in practice (and will not be known to RL algorithms before-hand either), it is easy to implement such a policy in simulation for comparative purposes.
We compare this rule-of-thumb to the optimal policy found by training an agent using the TD3 algorithm.

@@

Examining the replicate management trajectories and corresponding rewards [Fig 3] reveal that the RL agent incurs significantly higher costs in the initial phases of the simulation, dipping the cumulative reward into negative territory on average and well below the reward realized by the simple steady-state strategy.
This initial investment then begins to pay off -- by about 200th time step the RL agent has surpassed the performance of the steady state strategy.\
The policy plot provides more intuition for the RL agent's strategy: at very high state values, the RL agent opts for no conservation action -- so far from the tipping point, no response is required.
Near the tipping point, the RL agent steeply ramps up the conservation effort, and retains this effort even as the system falls below the critical threshold, where a sufficiently aggressive response can tip the system back into recovery.
For a system at or very close to the zero-state, the RL agent gives up, opting for no action.
Recall that the quadratic scaling of cost makes the rapid response of the TD3 algorithm much more costly to achieve the same net environmental improvement divided into smaller increments over a longer timeline.
However, our RL agent has discovered that the extra investment for a rapid response is well justified as the risk of crossing a tipping point increases.

@@

Such examination makes it clear how the RL agent's use of a more dynamic strategy out-preforms the steady-state strategy.
The various kinks in the RL policy function also suggest that this solution is nevertheless surely not the optimal solution -- such features are typical of RL-based policies, where such spandral features from early experimentation may become baked into the long-term strategy if adjusting them has little improvement on the final reward.
Further tuning of hyper-parameters, increased training, alterations to the training algorithm or an entirely different training algorithm would likely be able to further improve upon this performance.
The use of a recurrent neural network architecture instead of the multi-layer perceptron used here would open the door for the policy to reflect the non-autonomous dynamics more directly.
A recurrent neural network, such as long-short-term-memory (LSTM), could learn a strategy which reflects not only the current system state, but also latent information such as the time elapsed.
<!-- Kari: lots of jargon in this paragraph that needs to be defined-->

@@

-->
```
<!-- Something about relevance to both the _theory_ and practice of ecological conservation. (e.g. "strategic" 1-D models have probably had a greater influence, (for better or worse!) on real world policies than complex, detail-specific, precise "tactical" models) -->

@@

The examples presented here only scrape the surface of possible RL applications to conservation problems.
The examples we have focused on are intentionally quite simple, though it is worth remembering that these very same simple models have a long history of relevance and application in both research and policy contexts.
Despite their simplicity, the optimal strategy is not always obvious before hand, however intuitive it may appear in retrospect.
In the case of the ecosystem tipping point scenario, the optimal strategy is unknown, and the approximate solution found by our RL implementation could almost certainly be improved upon.

@@

-   wild west of AI?
    -can separate RL from having gyms implemented -\> how do you compare different models, -mention that customization is probably the way to go, baselines only go so far
